---
title: "Evaluate Final Classifiers"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
  word_document: null
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = T,
                      eval = T,
                      cache = F,
                      message = F,
                      warning = F)

HOME <- "C:/Users/Owner/OneDrive/Thesis/Thesis R Project"
DERIVED <- file.path(HOME, "Derived")
MODELS <- file.path(DERIVED, "Models")
WORKSPACES <- file.path(DERIVED, "WORKSPACES")
EXPORT <- file.path(HOME, "Export")
```

# Setup
```{r load.packages, include = F}
library(tidyr)
library(MASS)
library(gridExtra)
library(miscset)
library(PAA)
library(limma)
library(knitr)
library(doParallel)
library(Hmisc)
library(e1071)
library(randomForest)
library(kernlab)
library(caret)
library(dplyr)
library(reshape2)
library(ggplot2)
library(caTools)
```

```{r load.workspaces, include = F}
load(file = file.path(WORKSPACES,
                      "Modeling-Ready Workspace.RData"))

load(file = file.path(WORKSPACES,
                      "Custom Functions.RData"))
```

```{r load.models, include = F}
load(file = file.path(MODELS,
                      "finalRF.RData"))
load(file = file.path(MODELS,
                      "finalSVM.RData"))
```

```{r load.biomarkers, include = F}
load(file = file.path(DERIVED,
                      "biomarker.RF.RData"))
load(file = file.path(DERIVED,
                      "biomarker.SVM.RData"))
```

# Comparison of Resampling Performance
```{r}
resamp <- resamples(list(RF = finalRF, 
                         SVM = finalSVM))
modelDifferences <- diff(resamp)
```

95% confidence intervals of average resampled performance measures.
```{r}
resamp.RF <- 
  finalRF$resample %>%
  filter(mtry == 2) %>%
  select(-mtry, -Kappa) %>%
  mutate(Model = rep("RF", 500))
resamp.RF$Resample <- c(1:500)


resamp.SVM <-
  finalSVM$resample %>%
  filter(C == 0.25) %>%
  select(-C, -Kappa) %>%
  mutate(Model = rep("SVM", 500))

resamp.SVM$Resample <- c(1:500)

resampDF <- bind_rows(resamp.RF, resamp.SVM)
meltedResamp <- melt(resampDF,
                     id.vars = c("Resample", "Model"))
```

```{r}
tiff(filename = file.path(EXPORT, "Resamp Conf Intervals.tiff"),
     width = 2000,
     height = 2000,
     compression = "lzw",
     res = 300)

meltedResamp %>%
  group_by(variable, Model) %>%
  summarize(Mean = mean(value),
            SD = sd(value),
            LB = Mean - 1.96*(SD/sqrt(500)),
            UB = Mean + 1.96*(SD/sqrt(500))) %>%
  ggplot(aes(x = variable,
             y = Mean,
             color = factor(Model))) +
  geom_pointrange(aes(ymin = LB, ymax = UB)) +
  theme_bw() +
  coord_flip() +
  xlab("Evaluation Metric") +
  scale_color_brewer(name = "Model",
                     type = "qual",
                     palette = "Set1") +
  theme(axis.title = element_text(face = "bold"),
        axis.title.x = element_blank(),
        legend.title = element_text(face = "bold"))

dev.off()

```

Paired sample t-tests.
```{r}
OOBAUC.RF <- resamp.RF$AUC
OOBAUC.SVM <- resamp.SVM$AUC

OOBAcc.RF <- resamp.RF$Accuracy
OOBAcc.SVM <- resamp.SVM$Accuracy

OOBSens.RF <- resamp.RF$Sens
OOBSens.SVM <- resamp.SVM$Sens

OOBSpec.RF <- resamp.RF$Spec
OOBSpec.SVM <- resamp.SVM$Spec
```

AUC
```{r}
t.test(OOBAUC.RF, OOBAUC.SVM, paired = T)
```

Accuracy
```{r}
t.test(OOBAcc.RF, OOBAcc.SVM, paired = T)
```

Sensitivity
```{r}
t.test(OOBSens.RF, OOBSens.SVM, paired = T)
```

Specificity
```{r}
t.test(OOBSpec.RF, OOBSpec.SVM, paired = T)
```

# Evaluation on Test Set
## RF
Confusion matrix
```{r}
pred.RF <- predict(finalRF, 
                   data.test[, biomarker.RF])
confusionMatrix(pred.RF, 
                classVector.test, 
                positive = "AD")
```

```{r}
pred.prob.RF <- predict(finalRF, 
                        data.test[, biomarker.RF], 
                        type = "prob")
probs.RF <- pred.prob.RF
probs.RF$Sample <- rownames(probs.RF)
probs.RF$NDC <- NULL

probs.AD.RF <- probs.RF %>%
  filter(grepl(pattern = "^(AD)", 
               Sample))

probs.NDC.RF <- probs.RF %>%
  filter(!grepl(pattern = "^(AD)", 
                Sample))
```

prediction probabilities for AD samples
```{r, results = 'asis'}
probs.AD.RF %>%
  kable(digits = 4)
```

prediction probabilities for NDC samples
```{r, results = 'asis'}
probs.NDC.RF %>%
  kable(digits = 4)
```

ROC Curve Plots.
```{r}
df.RF <- data_frame(pred = pred.prob.RF$AD,
                    act = classVector.test)
roc.RF <- computeROC(df.RF, 
                     n = 1000)
```

```{r}
tiff(filename = file.path(EXPORT,
                          "RF ROC Plots.tiff"),
     width = 5.57,
     height = 2.81,
     units = "in",
     compression = "lzw",
     res = 200)

plots.roc(roc.RF, .5, "Random Forest")

dev.off()
```

Calculate test set AUC.
```{r}
trapz(roc.RF$fpr, 
      roc.RF$tpr)
```

## SVM
Confusion matrix
```{r}
pred.SVM <- predict(finalSVM, 
                    data.test[, biomarker.SVM])
confusionMatrix(pred.SVM, 
                classVector.test, 
                positive = "AD")
```

```{r}
pred.prob.SVM <- predict(finalSVM, 
                         data.test[, biomarker.SVM], 
                         type = "prob")
probs.SVM <- pred.prob.SVM
probs.SVM$Sample <- probs.RF$Sample
probs.SVM$NDC <- NULL

probs.AD.SVM <- probs.SVM %>%
  filter(grepl(pattern = "^(AD)", 
               Sample))

probs.NDC.SVM <- probs.SVM %>%
  filter(!grepl(pattern = "^(AD)", 
                Sample))
```

prediction probabilities for AD samples
```{r, results = 'asis'}
probs.AD.SVM %>%
  kable(digits = 4)
```

prediction probabilities for NDC samples
```{r, results = 'asis'}
probs.NDC.SVM %>%
  kable(digits = 4)
```

ROC Curve Plots.
```{r}
df.SVM <- data_frame(pred = pred.prob.SVM$AD,
                    act = classVector.test)
roc.SVM <- computeROC(df.SVM, 
                      n = 1000)
```

```{r}
tiff(filename = file.path(EXPORT,
                          "SVM ROC Plots.tiff"),
     width = 5.57,
     height = 2.81,
     units = "in",
     compression = "lzw",
     res = 200)

plots.roc(roc.SVM, .5, "Support Vector Machine")

dev.off()
```

Calculate test set AUC.
```{r}
trapz(roc.SVM$fpr, 
      roc.SVM$tpr)
```
