---
title: "Background Correction"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
  word_document: null
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = F,
                      eval = T,
                      cache = F,
                      message = F,
                      warning = F)

HOME <- "C:/Users/Owner/OneDrive/Thesis/Thesis R Project"
DERIVED <- file.path(HOME, "Derived")
MODELS <- file.path(DERIVED, "Models")
WORKSPACES <- file.path(DERIVED, "WORKSPACES")
EXPORT <- file.path(HOME, "Export")
```


```{r loadPackages, include = F}
library(tidyr)
library(MASS)
library(gridExtra)
library(miscset)
library(PAA)
library(limma)
library(knitr)
library(doParallel)
library(Hmisc)
library(e1071)
library(randomForest)
library(kernlab)
library(caret)
library(dplyr)
library(reshape2)
library(ggplot2)
library(DT)
```

# Setup
```{r importFiles}
load(file.path(WORKSPACES,
               "Data Preparation Output.RData"))
load(file.path(WORKSPACES,
               "Custom Functions.RData"))
load(file.path(DERIVED, "samples.MA.RData"))
```

# Background Correct Autoantibody and Control Data
```{r BC}
elist.data.BC <- backgroundCorrect(elist.data, 
                                   method = "subtract")
elist.controls.BC <- backgroundCorrect(elist.controls, 
                                       method = "subtract")
```

Melted autoantibody data
```{r}
data.BC <- as.data.frame(t(elist.data.BC$E))
data.BC$Sample <- rownames(data.BC)
rownames(data.BC) <- NULL

meltedData <- melt(data.BC,
                   id.vars = "Sample",
                   variable.name = "Feature")
meltedData <-
  meltedData %>%
  mutate(Class = if_else(grepl(pattern = "^(AD)", 
                               x = Sample, perl = T), 
                         "AD", 
                         "NDC"))
```

Melted control protein data
```{r}
controls.BC <- as.data.frame(t(elist.controls.BC$E))
controls.BC$Sample <- rownames(controls.BC)
rownames(controls.BC) <- NULL

meltedControls <- melt(controls.BC,
                       id.vars = "Sample",
                       variable.name = "Feature")
meltedControls <-
  meltedControls %>%
  mutate(Class = if_else(grepl(pattern = "^(AD)", 
                               x = Sample, perl = T), 
                         "AD", 
                         "NDC"))
```

Set any expression value less than 1 to 1.
```{r}
meltedData$value[which(meltedData$value < 1)] <- 1

meltedControls$value[which(meltedControls$value < 1)] <- 1
```

Log2 transform expression values.
```{r}
meltedData$value <- log2(meltedData$value)
meltedControls$value <- log2(meltedControls$value)
```

Eliminate autoantibodies with expression = 0 for all samples
```{r}
features.eliminated <-
  meltedData %>%
  group_by(Feature) %>%
  summarize(nonZeros = sum(value > 0)) %>%
  filter(nonZeros == 0) %>%
  select(Feature) %>%
  unlist()

meltedData <-
  meltedData %>%
  filter(Feature %nin% features.eliminated)

metaData.features <-
  metaData.features %>%
  filter(ID %nin% features.eliminated)
```

metaData of filtered autoantibodies
```{r metaData.features.eliminated, results = 'asis'}
metaData.features %>%
  filter(ID %in% features.eliminated) %>%
  select(ID, Description) %>%
  kable(format = "pandoc")
```

# Summary Statistics
## Autoantibodies
### Samples
```{r}
stats.samples <-
  meltedData %>%
  group_by(Sample) %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value),
            Outliers.high = sum(value > Q3 + (1.5 * IQR)),
            Outliers.low = sum(value < Q1 - (1.5 * IQR)))
```


```{r, results = 'asis'}
stats.samples %>%
  datatable() %>%
  formatRound(columns = colnames(stats.samples), digits = 4)
```

#### AD sample stats
```{r}
stats.samples.AD <- 
  meltedData %>%
  filter(Class == "AD") %>%
  group_by(Sample) %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value),
            Outliers.high = sum(value > Q3 + (1.5 * IQR)),
            Outliers.low = sum(value < Q1 - (1.5 * IQR)))
```

```{r, results = 'asis'}
stats.samples.AD %>%
  datatable() %>%
  formatRound(columns = colnames(stats.samples.AD), digits = 4)
```

#### NDC sample stats
```{r}
stats.samples.NDC <- 
  meltedData %>%
  filter(Class == "NDC") %>%
  group_by(Sample) %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value),
            Outliers.high = sum(value > Q3 + (1.5 * IQR)),
            Outliers.low = sum(value < Q1 - (1.5 * IQR)))
```

```{r, results = 'asis'}
stats.samples.NDC %>%
  datatable() %>%
  formatRound(columns = colnames(stats.samples.NDC), digits = 4)
```

#### Average sample stats
```{r}
aveStats.samples <-
  stats.samples %>%
  summarise(Min = mean(Min),
            Max = mean(Max),
            Mean = mean(Mean),
            Median = mean(Median),
            SD = mean(SD),
            Q1 = mean(Q1),
            Q3 = mean(Q3),
            IQR = mean(IQR),
            Outliers.high = mean(Outliers.high),
            Outliers.low = mean(Outliers.low))
```

```{r, results = 'asis'}
aveStats.samples %>%
  datatable() %>%
  formatRound(columns = colnames(aveStats.samples), digits = 4)
```

### Class
```{r}
stats.class <- 
  meltedData %>%
  group_by(Class) %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value))
```

```{r, results = 'asis'}
stats.class %>%
  datatable() %>%
  formatRound(columns = colnames(stats.class), digits = 4)
```

#### Average AD sample stats
```{r}
aveStats.samples.AD <-
  stats.samples.AD %>%
  summarise(Min = mean(Min),
            Max = mean(Max),
            Mean = mean(Mean),
            Median = mean(Median),
            SD = mean(SD),
            Q1 = mean(Q1),
            Q3 = mean(Q3),
            IQR = mean(IQR),
            Outliers.high = mean(Outliers.high),
            Outliers.low = mean(Outliers.low))
```

```{r, results = 'asis'}
aveStats.samples.AD %>%
  datatable() %>%
  formatRound(columns = colnames(aveStats.samples.AD), digits = 4)
```

#### Average NDC sample stats
```{r}
aveStats.samples.NDC <-
  stats.samples.NDC %>%
  summarise(Min = mean(Min),
            Max = mean(Max),
            Mean = mean(Mean),
            Median = mean(Median),
            SD = mean(SD),
            Q1 = mean(Q1),
            Q3 = mean(Q3),
            IQR = mean(IQR),
            Outliers.high = mean(Outliers.high),
            Outliers.low = mean(Outliers.low))
```

```{r, results = 'asis'}
aveStats.samples.NDC %>%
  datatable() %>%
  formatRound(columns = colnames(aveStats.samples.NDC), digits = 4)
```

### Variables
```{r}
stats.features <- 
  meltedData %>%
  group_by(Feature) %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value))

```

```{r, results = 'asis'}
stats.features %>%
  datatable() %>%
  formatRound(columns = colnames(stats.features), digits = 4)
```

#### Average autoantibody stats
```{r}
aveStats.features <-
  stats.features %>%
  summarise(Min = mean(Min),
            Max = mean(Max),
            Mean = mean(Mean),
            Median = mean(Median),
            SD = mean(SD),
            Q1 = mean(Q1),
            Q3 = mean(Q3),
            IQR = mean(IQR))
```

```{r, results = 'asis'}
aveStats.features %>%
  datatable() %>%
  formatRound(columns = colnames(aveStats.features), digits = 4)
```

#### SNR of each autoantibody:
```{r}
aveStats.features.perClass <- 
  meltedData %>%
  group_by(Feature, Class) %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value),
            Var = var(value))

SNR.features <-
  aveStats.features.perClass %>%
  group_by(Feature) %>%
  summarize(S = (Mean[Class == "AD"] - Mean[Class == "NDC"])^2 / 
              (Var[Class == "AD"] + Var[Class == "NDC"]))
```

```{r, results = 'asis'}
SNR.features %>%
  datatable() %>%
  formatRound(columns = colnames(SNR.features), digits = 4)
```

#### Distriubtion of Autoantibody SNRs
```{r}
summary(SNR.features$S)
```

### Summary stats of all data
```{r}
stats.signal <- 
  meltedData %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value))
```

```{r, results = 'asis'}
stats.signal %>%
  datatable() %>%
  formatRound(columns = colnames(stats.signal), digits = 4)
```

## Control Proteins
### Samples
```{r}
stats.samples.ctrls <-
  meltedControls %>%
  group_by(Sample) %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value),
            Outliers.high = sum(value > Q3 + (1.5 * IQR)),
            Outliers.low = sum(value < Q1 - (1.5 * IQR)))
```


```{r, results = 'asis'}
stats.samples.ctrls %>%
  datatable() %>%
  formatRound(columns = colnames(stats.samples.ctrls), digits = 4)
```

#### AD sample stats
```{r}
stats.samples.AD.ctrls <- 
  meltedControls %>%
  filter(Class == "AD") %>%
  group_by(Sample) %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value),
            Outliers.high = sum(value > Q3 + (1.5 * IQR)),
            Outliers.low = sum(value < Q1 - (1.5 * IQR)))
```

```{r, results = 'asis'}
stats.samples.AD.ctrls %>%
  datatable() %>%
  formatRound(columns = colnames(stats.samples.AD.ctrls), digits = 4)
```

#### NDC sample stats
```{r}
stats.samples.NDC.ctrls <- 
  meltedControls %>%
  filter(Class == "NDC") %>%
  group_by(Sample) %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value),
            Outliers.high = sum(value > Q3 + (1.5 * IQR)),
            Outliers.low = sum(value < Q1 - (1.5 * IQR)))
```

```{r, results = 'asis'}
stats.samples.NDC.ctrls %>%
  datatable() %>%
  formatRound(columns = colnames(stats.samples.NDC.ctrls), digits = 4)
```

#### Average sample stats
```{r}
aveStats.samples.ctrls <-
  stats.samples.ctrls %>%
  summarise(Min = mean(Min),
            Max = mean(Max),
            Mean = mean(Mean),
            Median = mean(Median),
            SD = mean(SD),
            Q1 = mean(Q1),
            Q3 = mean(Q3),
            IQR = mean(IQR),
            Outliers.high = mean(Outliers.high),
            Outliers.low = mean(Outliers.low))
```

```{r, results = 'asis'}
aveStats.samples.ctrls %>%
  datatable() %>%
  formatRound(columns = colnames(aveStats.samples.ctrls), digits = 4)
```

### Class
```{r}
stats.class.ctrls <- 
  meltedControls %>%
  group_by(Class) %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value))
```

```{r, results = 'asis'}
stats.class.ctrls %>%
  datatable() %>%
  formatRound(columns = colnames(stats.class.ctrls), digits = 4)
```

#### Average AD sample stats
```{r}
aveStats.samples.AD.ctrls <-
  stats.samples.AD.ctrls %>%
  summarise(Min = mean(Min),
            Max = mean(Max),
            Mean = mean(Mean),
            Median = mean(Median),
            SD = mean(SD),
            Q1 = mean(Q1),
            Q3 = mean(Q3),
            IQR = mean(IQR),
            Outliers.high = mean(Outliers.high),
            Outliers.low = mean(Outliers.low))
```

```{r, results = 'asis'}
aveStats.samples.AD.ctrls %>%
  datatable() %>%
  formatRound(columns = colnames(aveStats.samples.AD.ctrls), digits = 4)
```

#### Average NDC sample stats
```{r}
aveStats.samples.NDC.ctrls <-
  stats.samples.NDC.ctrls %>%
  summarise(Min = mean(Min),
            Max = mean(Max),
            Mean = mean(Mean),
            Median = mean(Median),
            SD = mean(SD),
            Q1 = mean(Q1),
            Q3 = mean(Q3),
            IQR = mean(IQR),
            Outliers.high = mean(Outliers.high),
            Outliers.low = mean(Outliers.low))
```

```{r, results = 'asis'}
aveStats.samples.NDC.ctrls %>%
  datatable() %>%
  formatRound(columns = colnames(aveStats.samples.NDC.ctrls), digits = 4)
```

### Variables
```{r}
stats.features.ctrls <- 
  meltedControls %>%
  group_by(Feature) %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value))

```

```{r, results = 'asis'}
stats.features.ctrls %>%
  datatable() %>%
  formatRound(columns = colnames(stats.features.ctrls), digits = 4)
```

#### Average control protein stats
```{r}
aveStats.features.ctrls <-
  stats.features.ctrls %>%
  summarise(Min = mean(Min),
            Max = mean(Max),
            Mean = mean(Mean),
            Median = mean(Median),
            SD = mean(SD),
            Q1 = mean(Q1),
            Q3 = mean(Q3),
            IQR = mean(IQR))
```

```{r, results = 'asis'}
aveStats.features.ctrls %>%
  datatable() %>%
  formatRound(columns = colnames(aveStats.features.ctrls), digits = 4)
```

### Summary stats of all data
```{r}
stats.signal.ctrls <- 
  meltedControls %>%
  summarise(Min = min(value),
            Max = max(value),
            Mean = mean(value),
            Median = median(value),
            SD = sd(value),
            Q1 = summary(value)[2],
            Q3 = summary(value)[5],
            IQR = IQR(value))
```

```{r, results = 'asis'}
stats.signal.ctrls %>%
  datatable() %>%
  formatRound(columns = colnames(stats.signal.ctrls), digits = 4)
```

# Array boxplots
Autoantibodies
```{r}
tiff(filename = file.path(EXPORT, "Array Boxplots BC.tiff"),
     width = 4000,
     height = 2000,
     pointsize = 10,
     compression = "lzw",
     res = 300)

meltedData %>%
  ggplot(aes(x = Sample,
             y = value,
             fill = Class)) +
  geom_boxplot(outlier.size = .5,
               outlier.alpha = .3,
               outlier.color = "darkgrey") +
  theme_bw() +
  scale_y_continuous(limits = c(0, 16),
                     breaks = seq(0, 16)) +
  scale_fill_brewer(palette = "Set1") +
  ylab(expression(bold(paste("Log"[bold(2)], " Expression")))) +
  theme(axis.text.x = element_text(size = 6,
                                   hjust = 1,
                                   angle = 45),
        axis.title.x = element_text(face = "bold"))
dev.off()

```

Controls
```{r}
tiff(filename = file.path(EXPORT, "Array Boxplots BC controls.tiff"),
     width = 4000,
     height = 2000,
     pointsize = 10,
     compression = "lzw",
     res = 300)

meltedControls %>%
  ggplot(aes(x = Sample,
             y = value,
             fill = Class)) +
  geom_boxplot(outlier.size = .5,
               outlier.alpha = .3,
               outlier.color = "darkgrey") +
  theme_bw() +
  scale_y_continuous(limits = c(0, 16),
                     breaks = seq(0, 16)) +
  scale_fill_brewer(palette = "Set1") +
  ylab(expression(bold(paste("Log"[bold(2)], " Expression")))) +
  theme(axis.text.x = element_text(size = 6,
                                   hjust = 1,
                                   angle = 45),
        axis.title.x = element_text(face = "bold"))
dev.off()
```

# MA plots
```{r plot.MAs}
plot.MAs <- function(data.melted,
                     value.var,
                     need.to.log2 = F,
                     samples) {
  n.samples <- length(samples)
  plotList <- list()
  for (i in 1:n.samples){
    sample <- samples[i]
    plotList[[sample]] <- 
      plot.MA(data.melted = data.melted,
              array = sample,
              value.var = value.var,
              need.to.log2 = need.to.log2) +
      scale_x_continuous(limits = c(0, 16),
                         breaks = seq(0, 16, 
                                      by = 2)) +
      scale_y_continuous(limits = c(-14, 14),
                         breaks = seq(-14, 14,
                                      by = 2))
  }
  return(plotList)
}
```

```{r}
list.MAPlots <- plot.MAs(data.melt = meltedData,
                         value.var = "value",
                         need.to.log2 = F,
                         samples = samples.MA)
```

```{r}
tiff(filename = file.path(EXPORT, "MAPlots.BC.tiff"),
     width = 4000,
     height = 4000,
     pointsize = 25,
     compression = "lzw",
     res = 300)

ggplotGrid(list.MAPlots,
           ncol = 3,
           nrow = 3)

dev.off()
```

# Save
```{r Save}
meltedData$Class <- NULL
meltedControls$Class <- NULL

save(meltedData,
     meltedControls,
     metaData.samples,
     metaData.features,
     metaData.controls,
     file = file.path(WORKSPACES,
                      "BC Output.RData"))
```