---
title: "RLM Normalization"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
  word_document: null
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = F,
                      eval = T,
                      cache = F,
                      message = F,
                      warning = F)

HOME <- "C:/Users/Owner/OneDrive/Thesis/Thesis R Project"
DERIVED <- file.path(HOME, "Derived")
MODELS <- file.path(DERIVED, "Models")
WORKSPACES <- file.path(DERIVED, "WORKSPACES")
EXPORT <- file.path(HOME, "Export")
```

```{r loadPackages, include = F}
library(tidyr)
library(MASS)
library(gridExtra)
library(miscset)
library(PAA)
library(limma)
library(knitr)
library(doParallel)
library(Hmisc)
library(e1071)
library(randomForest)
library(kernlab)
library(caret)
library(dplyr)
library(reshape2)
library(ggplot2)
library(DT)
```


# Setup
```{r importFiles}
load(file.path(WORKSPACES,
               "BC Output.RData"))
load(file.path(WORKSPACES,
               "Custom Functions.RData"))
load(file.path(DERIVED, "samples.MA.RData"))
```

Add Block and random error columns to melted controls.
```{r}
meltedControls$Block <- rep(metaData.controls$Block,
                            times = length(metaData.samples$ArrayID))

meltedControls$e <- rnorm(length(meltedControls$value), 
                          mean = 0,
                          sd = 1)

meltedControls$Sample <- as.factor(meltedControls$Sample)
meltedControls$Feature <- as.factor(meltedControls$Feature)
meltedControls$Block <- as.factor(meltedControls$Block)
```

Make dummy variables.
```{r}
dummies <- dummyVars(value ~ ., 
                     data = meltedControls,
                     fullRank = T)
meltedControls.dummies <- predict(dummies, 
                                  newdata = meltedControls)
meltedControls.dummies <- as.data.frame(meltedControls.dummies)
meltedControls.dummies$value <- meltedControls$value
```

# RLM Normalization
```{r}
rlm.train <- rlm(value ~ .,
                 data = meltedControls.dummies)
```

```{r}
coefs <- as.data.frame(coef(rlm.train))
colnames(coefs) <- c("Value")
coefs$Coefficient <- rownames(coefs)
rownames(coefs) <- NULL
coefs <- coefs[, c(2, 1)]
```

```{r}
coefs.samples <-
  coefs %>%
  filter(grepl(pattern = "^(Sample)", 
               Coefficient))
coefs.blocks <-
  coefs %>%
  filter(grepl(pattern = "^(Block)", 
               Coefficient))
```

```{r}
coefs.samples$Coefficient <- gsub(pattern = "Sample.", 
                                  replacement = "", 
                                  x = coefs.samples$Coefficient)

coefs.blocks$Coefficient <- gsub(pattern = "Block.", 
                                 replacement = "", 
                                 x = coefs.blocks$Coefficient)
```

```{r}
coefs.samples <- rbind(coefs.samples,
                       c("Coefficient" = "AD1",
                         "Value" = 0))

coefs.blocks <- rbind(coefs.blocks,
                      c("Coefficient" = "1",
                        "Value" = 0))
```

```{r}
coefDF.samples <- coefs.samples
coefDF.blocks <- coefs.blocks
colnames(coefDF.samples) <- c("Sample", 
                              "Coefficient")
colnames(coefDF.blocks) <- c("Block", 
                             "Coefficient")
```

Sample coefficients:
```{r, results = 'asis'}
kable(coefDF.samples, 
      format = "pandoc")
```

Block coefficients:
```{r, results = 'asis'}
kable(coefDF.blocks, 
      format = "pandoc")
```

```{r}
meltedData.RLM <- meltedData
meltedData.RLM$Block <- 
  rep(metaData.features$Block,
                            times = length(metaData.samples$ArrayID)) %>%
  as.factor()
```

```{r}
meltedData.RLM$Coef.Block <-
  meltedData.RLM$Block %>%
  sapply(FUN = function(x){
    coefs.blocks$Value[which(coefs.blocks$Coefficient == x)]
  })

meltedData.RLM$Coef.Sample <-
  meltedData.RLM$Sample %>%
  sapply(FUN = function(x){
    coefs.samples$Value[which(coefs.samples$Coefficient == x)]
  })
```

```{r}
meltedData.RLM$value <- as.numeric(meltedData.RLM$value)
meltedData.RLM$Coef.Sample <- as.numeric(meltedData.RLM$Coef.Sample)
meltedData.RLM$Coef.Block <- as.numeric(meltedData.RLM$Coef.Block)
```

RLM normalized expression is original value minus the sample and block coefficients.
```{r}
meltedData.RLM <-
  meltedData.RLM %>%
  mutate(value.normalized = value - Coef.Sample - Coef.Block)
```

Convert melted normalized data back to wide dataframe format.
```{r}
data.all <- dcast(meltedData.RLM, 
                  formula = Sample ~ Feature, 
                  value.var = "value.normalized")
rownames(data.all) <- data.all$Sample
data.all$Sample <- NULL
```

```{r}
metaData.samples <- 
  metaData.samples %>%
  arrange(ArrayID, 
          match(metaData.samples$ArrayID, 
                rownames(data.all)))

metaData.features <-
  metaData.features %>%
  arrange(ID, 
          match(metaData.features$ID,
                colnames(data.all)))
```

# Summary Statistics
```{r}
meltedData.RLM <-
  meltedData.RLM %>%
  mutate(Class = if_else(grepl(pattern = "^(AD)", 
                               x = Sample, perl = T), 
                         "AD", 
                         "NDC"))
```

## Samples
```{r}
stats.samples <-
  meltedData.RLM %>%
  group_by(Sample) %>%
  summarise(Min = min(value.normalized),
            Max = max(value.normalized),
            Mean = mean(value.normalized),
            Median = median(value.normalized),
            SD = sd(value.normalized),
            Q1 = summary(value.normalized)[2],
            Q3 = summary(value.normalized)[5],
            IQR = IQR(value.normalized),
            Outliers.high = sum(value.normalized > Q3 + (1.5 * IQR)),
            Outliers.low = sum(value.normalized < Q1 - (1.5 * IQR)))
```


```{r, results = 'asis'}
stats.samples %>%
  datatable() %>%
  formatRound(columns = colnames(stats.samples), digits = 4)
```

### AD sample stats
```{r}
stats.samples.AD <- 
  meltedData.RLM %>%
  filter(Class == "AD") %>%
  group_by(Sample) %>%
  summarise(Min = min(value.normalized),
            Max = max(value.normalized),
            Mean = mean(value.normalized),
            Median = median(value.normalized),
            SD = sd(value.normalized),
            Q1 = summary(value.normalized)[2],
            Q3 = summary(value.normalized)[5],
            IQR = IQR(value.normalized),
            Outliers.high = sum(value.normalized > Q3 + (1.5 * IQR)),
            Outliers.low = sum(value.normalized < Q1 - (1.5 * IQR)))
```

```{r, results = 'asis'}
stats.samples.AD %>%
  datatable() %>%
  formatRound(columns = colnames(stats.samples.AD), digits = 4)
```

### NDC sample stats
```{r}
stats.samples.NDC <- 
  meltedData.RLM %>%
  filter(Class == "NDC") %>%
  group_by(Sample) %>%
  summarise(Min = min(value.normalized),
            Max = max(value.normalized),
            Mean = mean(value.normalized),
            Median = median(value.normalized),
            SD = sd(value.normalized),
            Q1 = summary(value.normalized)[2],
            Q3 = summary(value.normalized)[5],
            IQR = IQR(value.normalized),
            Outliers.high = sum(value.normalized > Q3 + (1.5 * IQR)),
            Outliers.low = sum(value.normalized < Q1 - (1.5 * IQR)))
```

```{r, results = 'asis'}
stats.samples.NDC %>%
  datatable() %>%
  formatRound(columns = colnames(stats.samples.NDC), digits = 4)
```

### Average sample stats
```{r}
aveStats.samples <-
  stats.samples %>%
  summarise(Min = mean(Min),
            Max = mean(Max),
            Mean = mean(Mean),
            Median = mean(Median),
            SD = mean(SD),
            Q1 = mean(Q1),
            Q3 = mean(Q3),
            IQR = mean(IQR),
            Outliers.high = mean(Outliers.high),
            Outliers.low = mean(Outliers.low))
```

```{r, results = 'asis'}
aveStats.samples %>%
  datatable() %>%
  formatRound(columns = colnames(aveStats.samples), digits = 4)
```

## Class
```{r}
stats.class <- 
  meltedData.RLM %>%
  group_by(Class) %>%
  summarise(Min = min(value.normalized),
            Max = max(value.normalized),
            Mean = mean(value.normalized),
            Median = median(value.normalized),
            SD = sd(value.normalized),
            Q1 = summary(value.normalized)[2],
            Q3 = summary(value.normalized)[5],
            IQR = IQR(value.normalized))
```

```{r, results = 'asis'}
stats.class %>%
  datatable() %>%
  formatRound(columns = colnames(stats.class), digits = 4)
```

### Average AD sample stats
```{r}
aveStats.samples.AD <-
  stats.samples.AD %>%
  summarise(Min = mean(Min),
            Max = mean(Max),
            Mean = mean(Mean),
            Median = mean(Median),
            SD = mean(SD),
            Q1 = mean(Q1),
            Q3 = mean(Q3),
            IQR = mean(IQR),
            Outliers.high = mean(Outliers.high),
            Outliers.low = mean(Outliers.low))
```

```{r, results = 'asis'}
aveStats.samples.AD %>%
  datatable() %>%
  formatRound(columns = colnames(aveStats.samples.AD), digits = 4)
```

### Average NDC sample stats
```{r}
aveStats.samples.NDC <-
  stats.samples.NDC %>%
  summarise(Min = mean(Min),
            Max = mean(Max),
            Mean = mean(Mean),
            Median = mean(Median),
            SD = mean(SD),
            Q1 = mean(Q1),
            Q3 = mean(Q3),
            IQR = mean(IQR),
            Outliers.high = mean(Outliers.high),
            Outliers.low = mean(Outliers.low))
```

```{r, results = 'asis'}
aveStats.samples.NDC %>%
  datatable() %>%
  formatRound(columns = colnames(aveStats.samples.NDC), digits = 4)
```

## Variables
```{r}
stats.features <- 
  meltedData.RLM %>%
  group_by(Feature) %>%
  summarise(Min = min(value.normalized),
            Max = max(value.normalized),
            Mean = mean(value.normalized),
            Median = median(value.normalized),
            SD = sd(value.normalized),
            Q1 = summary(value.normalized)[2],
            Q3 = summary(value.normalized)[5],
            IQR = IQR(value.normalized))

```

```{r, results = 'asis'}
stats.features %>%
  datatable() %>%
  formatRound(columns = colnames(stats.features), digits = 4)
```

### Average autoantibody stats
```{r}
aveStats.features <-
  stats.features %>%
  summarise(Min = mean(Min),
            Max = mean(Max),
            Mean = mean(Mean),
            Median = mean(Median),
            SD = mean(SD),
            Q1 = mean(Q1),
            Q3 = mean(Q3),
            IQR = mean(IQR))
```

```{r, results = 'asis'}
aveStats.features %>%
  datatable() %>%
  formatRound(columns = colnames(aveStats.features), digits = 4)
```

### SNR of each autoantibody:
```{r}
aveStats.features.perClass <- 
  meltedData.RLM %>%
  group_by(Feature, Class) %>%
  summarise(Min = min(value.normalized),
            Max = max(value.normalized),
            Mean = mean(value.normalized),
            Median = median(value.normalized),
            SD = sd(value.normalized),
            Q1 = summary(value.normalized)[2],
            Q3 = summary(value.normalized)[5],
            IQR = IQR(value.normalized),
            Var = var(value.normalized))

SNR.features <-
  aveStats.features.perClass %>%
  group_by(Feature) %>%
  summarize(S = (Mean[Class == "AD"] - Mean[Class == "NDC"])^2 / 
              (Var[Class == "AD"] + Var[Class == "NDC"]))
```

```{r, results = 'asis'}
SNR.features %>%
  datatable() %>%
  formatRound(columns = colnames(SNR.features), digits = 4)
```

### Distriubtion of Autoantibody SNRs
```{r}
summary(SNR.features$S)
```

## Summary stats of all data
```{r}
stats.signal <- 
  meltedData.RLM %>%
  summarise(Min = min(value.normalized),
            Max = max(value.normalized),
            Mean = mean(value.normalized),
            Median = median(value.normalized),
            SD = sd(value.normalized),
            Q1 = summary(value.normalized)[2],
            Q3 = summary(value.normalized)[5],
            IQR = IQR(value.normalized))
```

```{r, results = 'asis'}
stats.signal %>%
  datatable() %>%
  formatRound(columns = colnames(stats.signal), digits = 4)
```

# Array boxplots
```{r}
tiff(filename = file.path(EXPORT, "Array Boxplots RLM.tiff"),
     width = 4000,
     height = 2000,
     pointsize = 10,
     compression = "lzw",
     res = 300)

meltedData.RLM %>%
  ggplot(aes(x = Sample,
             y = value.normalized,
             fill = Class)) +
  geom_boxplot(outlier.size = .5,
               outlier.alpha = .3,
               outlier.color = "darkgrey") +
  theme_bw() +
  scale_y_continuous(limits = c(0, 16),
                     breaks = seq(0, 16)) +
  scale_fill_brewer(palette = "Set1") +
  ylab(expression(bold(paste("Log"[bold(2)], " Expression")))) +
  theme(axis.text.x = element_text(size = 6,
                                   hjust = 1,
                                   angle = 45),
        axis.title.x = element_text(face = "bold"))
dev.off()
```

# MA plots
```{r plot.MAs}
plot.MAs <- function(data.melted,
                     value.var,
                     need.to.log2 = F,
                     samples) {
  n.samples <- length(samples)
  plotList <- list()
  for (i in 1:n.samples){
    sample <- samples[i]
    plotList[[sample]] <- 
      plot.MA(data.melted = data.melted,
              array = sample,
              value.var = value.var,
              need.to.log2 = need.to.log2) +
      scale_x_continuous(limits = c(0, 16),
                         breaks = seq(0, 16, 
                                      by = 2)) +
      scale_y_continuous(limits = c(-14, 14),
                         breaks = seq(-14, 14,
                                      by = 2))
  }
  return(plotList)
}
```

```{r}
list.MAPlots <- plot.MAs(data.melt = meltedData.RLM,
                         value.var = "value.normalized",
                         need.to.log2 = F,
                         samples = samples.MA)
```

```{r}
tiff(filename = file.path(EXPORT, "MAPlots.RLM.tiff"),
         width = 4000,
         height = 4000,
         pointsize = 25,
         compression = "lzw",
         res = 300)

    ggplotGrid(list.MAPlots,
           ncol = 3,
           nrow = 3)
    
    dev.off()
```

# Save
```{r Save}
save(data.all,
     metaData.samples,
     metaData.features,
     file = file.path(WORKSPACES,
                      "RLM Output.RData"))
```
