---
title: "RFE Results, Extract Biomarkers, Tune Final Classifiers"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
  word_document: null
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = T,
                      eval = T,
                      cache = F,
                      message = F,
                      warning = F)

HOME <- "C:/Users/Owner/OneDrive/Thesis/Thesis R Project"
DERIVED <- file.path(HOME, "Derived")
MODELS <- file.path(DERIVED, "Models")
WORKSPACES <- file.path(DERIVED, "WORKSPACES")
EXPORT <- file.path(HOME, "Export")
```

```{r load.packages, include = F}
library(tidyr)
library(MASS)
library(gridExtra)
library(miscset)
library(PAA)
library(limma)
library(knitr)
library(doParallel)
library(Hmisc)
library(e1071)
library(randomForest)
library(kernlab)
library(caret)
library(dplyr)
library(reshape2)
library(ggplot2)
library(DT)
```

```{r load.workspaces, include = F}
# Modeling-Ready Workspace
load(file = file.path(WORKSPACES,
                      "Modeling-Ready Workspace.RData"))

# Custom Functions for Train and RFE
load(file = file.path(WORKSPACES,
                      "Custom Functions.RData"))

# Load rf.RFE model
load(file = file.path(MODELS,
                      "rf.RFE.RData"))

# Load svm.RFE model
load(file = file.path(MODELS,
                      "svm.RFE.RData"))
```

# Setup
Change ROC labels to AUC.
```{r}
colnames(rf.RFE$results)[2] <- "AUC"
colnames(rf.RFE$results)[7] <- "AUCSD"

colnames(svm.RFE$results)[2] <- "AUC"
colnames(svm.RFE$results)[7] <- "AUCSD"


colnames(rf.RFE$resample)[2] <- "AUC"
colnames(svm.RFE$resample)[2] <- "AUC"
```

# RF
## RFE results
Average performance measures among the 500 bootstrapped RF models per subset size.
```{r, results = 'asis'}
kable(rf.RFE$results[, c(1:6)], 
      digits = 4,
      format = "pandoc")
```

Plot of average performance per subset size.
```{r}
results.RF <- rf.RFE$results[, c(1:3, 5, 6)]

meltedResults.RF <- melt(results.RF,
                         id.vars = "Variables",
                         variable.name = "Metric")
```

```{r}
tiff(filename = file.path(EXPORT,
                          "RF Eval Metric vs FSS.tiff"),
     width = 2000,
     height = 2000,
     compression = "lzw",
     res = 300)

meltedResults.RF %>%
  ggplot(aes(x = rep(c(1:38), 4),
             y = value,
             color = Metric)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  xlab("Subset Size")  +
  scale_x_continuous(breaks = c(1:38),
                     labels = FSS.rerank) +
  scale_color_brewer(name = "Evaluation Metric",
                     type = "qual",
                     palette = "Set1") +
  theme(axis.title = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        axis.title.x = element_blank()) +
  coord_flip()

dev.off()
```

Strip plot of AUC per subset size.
```{r}
resampPerf.RF <- rf.RFE$resample
resampPerf.RF$Variables <- as.factor(resampPerf.RF$Variables)
```

```{r}
tiff(filename = file.path(EXPORT,
                          "RF strip plot AUC vs FSS.tiff"),
     width = 2000,
     height = 2000,
     compression = "lzw",
     res = 300)
resampPerf.RF %>%
  select(1:3, 5, 6) %>%
  group_by(Variables) %>%
  ggplot(aes(x = AUC,
             y = Variables)) +
  geom_point(size = .7,
             color = "darkblue") +
  theme_bw() +
  scale_x_continuous(breaks = seq(.7, 1, by = .05)) +
  theme(axis.title = element_text(face = "bold"),
        axis.text.y = element_text(size = 8))
dev.off()
```

Summary stats of resampling results
```{r}
var.AUC.RF <- 
  resampPerf.RF %>%
  select(1, 2) %>%
  group_by(Variables) %>%
  summarise(var = var(AUC))
```


## Extract biomarker.
Determine optimal subset size.
```{r}
newSize.RF <- rf.RFE$results[tolerance(rf.RFE$results, "AUC", 
                                       maximize = TRUE, 
                                       tol = 1),
                             "Variables"]
biomarker.RF <- rf.functions$selectVar(rf.RFE$variables, 
                                       size = newSize.RF)
```

Export RF biomarker training set to .csv for clustering.
```{r}
tempDF <- data.training[, biomarker.RF]
colnames(tempDF) <-
  colnames(tempDF) %>%
  plyr::revalue(c("B07R02C17" = "PRELID2",
                  "B18R07C13" = "DNAJC8",
                  "B24R17C03" = "PCBD2",
                  "B36R03C09" = "MGC24125",
                  "B26R16C03" = "IL20",
                  "B17R09C11" = "FRMD8"))
write.csv(tempDF, 
          file = file.path(EXPORT,
                           "biomarker.RF.csv"),
          row.names = T,
          col.names = T)
```

RF biomarker metadata.
```{r}
metaData.RF <-
  metaData.features %>%
  filter(ID %in% biomarker.RF) %>%
  select(c(Description, ID))
metaData.RF <- metaData.RF[, c(2,1)]

metaData.RF$Feature <- 
  metaData.RF$ID %>%
  plyr::revalue(c("B07R02C17" = "PRELID2",
                  "B18R07C13" = "DNAJC8",
                  "B24R17C03" = "PCBD2",
                  "B36R03C09" = "MGC24125",
                  "B26R16C03" = "IL20",
                  "B17R09C11" = "FRMD8"))
metaData.RF <- metaData.RF[, c(3, 1, 2)]
```

```{r, results = 'asis'}
metaData.RF %>%
  kable()
```


## Final classifier
Parameter tuning and final model fit.
```{r}
cl <- makeCluster(detectCores())
registerDoParallel(cl)

set.seed(10)
finalRF <- train(x = as.matrix(data.training[, biomarker.RF]),
                 y = classVector.training,
                 method = "rf",
                 metric = "ROC",
                 tuneLength = 5,
                 trControl = trCtrl,
                 ntree = 2000,
                 importance = T,
                 keep.inbag = T)
stopCluster(cl)

colnames(finalRF$results)[2] <- "AUC"
colnames(finalRF$results)[7] <- "AUCSD"
colnames(finalRF$resample)[1] <- "AUC"
finalRF$perfNames[1] <- "AUC"
```

```{r}
save(finalRF,
     file = file.path(MODELS,
                      "finalRF.RData"))

save(biomarker.RF,
     file = file.path(DERIVED,
                      "biomarker.RF.RData"))
```

```{r, include = F}
load(file = file.path(MODELS,
                      "finalRF.RData"))
```

Tuning parameter performance.
```{r, results = 'asis'}
kable(finalRF$results[, c(1:6)], 
      digits = 4,
      format = "pandoc")
```

Internal OOB training performance of final RF model.
```{r}
finalRF$finalModel
```

Variable importance rankings.
```{r, results = 'asis'}
vImp.RF <- 
  rf.functions$rank(finalRF$finalModel)
colnames(vImp.RF)[1] <- "ID"

vImp.RF <-
  vImp.RF %>%
  left_join(metaData.RF[, c("ID", "Feature")], 
            by = "ID") %>%
  select(7, 1, 5, 3, 4, 6)
rownames(vImp.RF) <- NULL
vImp.RF$Feature <- factor(vImp.RF$Feature, 
                          levels = vImp.RF$Feature[order(vImp.RF$MeanDecreaseAccuracy)])

vImp.RF %>%
  kable(digits = 4,
        format = "pandoc")
```

Variable importance plot.
```{r}
tiff(filename = file.path(EXPORT,
                          "RF varImp plot.tiff"),
     width = 1554,
     height = 1159,
     compression = "lzw",
     res = 300)

vImp.RF %>%
  ggplot(aes(x = MeanDecreaseAccuracy,
             y = Feature)) +
  geom_point() +
  theme_bw() +
  xlab("Mean Decrease in Accuracy") +
  ylab("Feature") +
  theme(axis.title = element_text(face = "bold"))

dev.off()
```

# SVM
## RFE results
Average performance measures among the 500 bootstrapped SVM models per subset size.
```{r, results = 'asis'}
kable(svm.RFE$results[, c(1:6)], 
      digits = 4,
      format = "pandoc")
```

Plot of average performance per subset size.
```{r}
tiff(filename = file.path(EXPORT,
                          "SVM Eval Metric vs FSS.tiff"),
     width = 2000,
     height = 2000,
     compression = "lzw",
     res = 300)

results.SVM <- svm.RFE$results[, c(1:3, 5, 6)]

meltedResults.SVM <- melt(results.SVM,
                          id.vars = "Variables",
                          variable.name = "Metric")

meltedResults.SVM %>%
  ggplot(aes(x = rep(c(1:38), 4),
             y = value,
             color = Metric)) +
  geom_point() +
  geom_line() +
  theme_bw() +
  xlab("# Variables")  +
  scale_x_continuous(breaks = c(1:38),
                     labels = FSS.rerank) +
  scale_color_brewer(name = "Evaluation Metric",
                     type = "qual",
                     palette = "Set1") +
  theme(axis.title = element_text(face = "bold"),
        legend.title = element_text(face = "bold"),
        axis.title.x = element_blank()) +
  coord_flip()

dev.off()
```

```{r}
resampPerf.SVM <- svm.RFE$resample
resampPerf.SVM$Variables <- as.factor(resampPerf.SVM$Variables)
```

Strip plot of AUC per subset size.
```{r}
tiff(filename = file.path(EXPORT,
                          "SVM strip plot AUC vs FSS.tiff"),
     width = 2000,
     height = 2000,
     compression = "lzw",
     res = 300)

resampPerf.SVM %>%
  select(1:3, 5, 6) %>%
  group_by(Variables) %>%
  ggplot(aes(x = AUC,
             y = Variables)) +
  geom_point(size = .7,
             color = "darkblue") +
  theme_bw() +
  scale_x_continuous(breaks = seq(.7, 1, by = .05)) +
  theme(axis.title = element_text(face = "bold"),
        axis.text.y = element_text(size = 8))
dev.off()
```

Summary stats of resampling results
```{r}
var.AUC.SVM <- 
  resampPerf.SVM %>%
  select(1, 2) %>%
  group_by(Variables) %>%
  summarise(var = var(AUC))
```

```{r, results = 'asis'}
var.AUC.SVM %>%
  datatable()
```


## Extract Biomarker
```{r}
newSize.SVM <- svm.RFE$results[tolerance(svm.RFE$results[1:29, ], 
                                         "AUC",
                                         maximize = TRUE, 
                                         tol = 1),
                               "Variables"]
biomarker.SVM <- svm.functions$selectVar(svm.RFE$variables, 
                                         size = newSize.SVM)
```

Export SVM biomarker training set to .csv for clustering.
```{r}
tempDF <- data.training[, biomarker.SVM]
colnames(tempDF) <-
  colnames(tempDF) %>%
  plyr::revalue(c("B02R19C01" = "GTF2I",
                  "B35R11C21" = "LGALS1",
                  "B17R09C11" = "FRMD8",
                  "B21R12C15" = "PHF15",
                  "B24R17C03" = "PCBD2",
                  "B28R16C01" = "IL4",
                  "B25R11C21" = "PTCD2",
                  "B26R16C03" = "IL20",
                  "B15R12C09" = "ANKHD1"))
write.csv(tempDF, 
          file = file.path(EXPORT,
                           "biomarker.SVM.csv"),
          row.names = T,
          col.names = T)
```

SVM biomarker metadata.
```{r}
metaData.SVM <-
  metaData.features %>%
  filter(ID %in% biomarker.SVM) %>%
  dplyr::select(c(Description, ID))
metaData.SVM <- metaData.SVM[, c(2,1)]

metaData.SVM$Feature <- 
  metaData.SVM$ID %>%
  plyr::revalue(c("B02R19C01" = "GTF2I",
                  "B35R11C21" = "LGALS1",
                  "B17R09C11" = "FRMD8",
                  "B21R12C15" = "PHF15",
                  "B24R17C03" = "PCBD2",
                  "B28R16C01" = "IL4",
                  "B25R11C21" = "PTCD2",
                  "B26R16C03" = "IL20",
                  "B15R12C09" = "ANKHD1"))
metaData.SVM <- metaData.SVM[, c(3, 1, 2)]
```

## Final Classifier
Parameter tuning and final model fit.
```{r}
cl <- makeCluster(detectCores())
registerDoParallel(cl)

set.seed(10)
finalSVM <- train(x = as.matrix(data.training[, biomarker.SVM]),
                  y = classVector.training,
                  method = "svmLinear",
                  metric = "ROC",
                  tuneGrid = expand.grid(C = 2^c(-2:2)),
                  trControl = trCtrl)
stopCluster(cl)

colnames(finalSVM$results)[2] <- "AUC"
colnames(finalSVM$results)[7] <- "AUCSD"
colnames(finalSVM$resample)[1] <- "AUC"
finalSVM$perfNames[1] <- "AUC"
```

```{r}
save(finalSVM,
     file = file.path(MODELS,
                      "finalSVM.RData"))
save(biomarker.SVM,
     file = file.path(DERIVED,
                      "biomarker.SVM.RData"))
```

Tuning parameter performance.
```{r, results = 'asis'}
kable(finalSVM$results[, c(1:6)], 
      digits = 4,
      format = "pandoc")
```

Final SVM Details.
```{r}
finalSVM$finalModel
```

Variable importance rankings.
```{r, results = 'asis'}
vImp.SVM <- svm.functions$rank(finalSVM)
colnames(vImp.SVM)[1] <- "ID"
vImp.SVM <-
  vImp.SVM %>%
  left_join(metaData.SVM[, c("ID", 
                             "Feature")],
            by = "ID") %>%
  select(3, 1, 2)
rownames(vImp.SVM) <- NULL
vImp.SVM$Feature <- factor(vImp.SVM$Feature,
                           levels = vImp.SVM$Feature[order(vImp.SVM$Overall)])

vImp.SVM %>%
  kable(digits = 4,
        format = "pandoc")
```

Variable importance plot.
```{r}
tiff(filename = file.path(EXPORT,
                          "SVM varImp plot.tiff"),
     width = 1554,
     height = 1159,
     compression = "lzw",
     res = 300)

vImp.SVM %>%
  ggplot(aes(x = Overall,
             y = Feature)) +
  geom_point() +
  theme_bw() +
  xlab("Influence on Weight Vector") +
  ylab("Feature") +
  theme(axis.title = element_text(face = "bold"))
dev.off()

```